---
title: "SOTU Analysis - POLI 176 Final Project"
format: html
editor: visual
---

```{r}
library(dplyr) 
library(tidyverse)
library(tokenizers)
library(quanteda)
library(quanteda.textmodels)
library(caret)
library(irr)
```

# 1.

```{r}
sotu <- read.csv("SOTU_WithText.csv")

# filter from 1900s on 
filtered_sotu <- filter(sotu, year >= 1900)

```

Select a corpus to study. Do some background reading on the corpus and decide on a research question that you can answer with the data that you have. What is your research question? What is the significance of your research question? Do a quick search -- has anyone studied this question before? (Please include at least 2 references) \[2-3 paragraphs\]

The State of the Union Address (SOTU) is a communication from the President to Congress, typically in written or verbal form (History, Evolution, and Practices of the President’s State of the Union Address: Frequently Asked Questions, n.d.). State of the Union Addresses are a chance for the President to report and share on the general state of affairs in the country - important and pressing issues are usually emphasized, in addition to legislative actions the President recommends to rectify them to guide the country towards their vision of the future. Thus, these speeches offer a glimpse into the sentiments held by Presidents, the constituents and politicians of the political party they represent, and the broader feelings of the American public who voted them into office. With the recent threats to Medicare, Medicaid, and the Affordable Care Act, we were interested in exploring how sentiment regarding public health care has shifted over time and how political party affiliation affects it (Copeland, 2025). Thus, our research question for this project is:

Are there systematic differences in how Democratic and Republican presidents frame public health issues in terms of sentiment (approval vs. neutral vs. disapproval) in State of the Union Addresses?

Doing a quick search, we were unable to find any articles or literature that address our same question; however, there are related works that address different aspects of the research question. For example, an article titled "Health Features Prominently in the State of the Union Address" by Terri Rutter briefly discusses President Clinton's 1998 State of the Union Address, and his discourse and sentiment surrounding public health and healthcare (Rutter). Further, another article by Ryan L. Teten titled "Evolution of the Modern Rhetorical Presidency: Presidential Presentation and Development of the State of the Union Address" discusses in more detail how the sentiment in State of the Union addresses have developed over time, but doesn't compare differences in sentiment between political parties (Teten).

For this project, we created 3 categories of sentiment in State of the Union speeches regarding public health care: approval, neutral, and disapproval. Our definitions are as follows:

```         
-   **Approval**: SOTU speech expressing satisfaction or belief in the value of a public health program or initiative. This can also take the form of having disapproval for private health initiatives (i.e. lack of support of the privatization of health care programs).
-   **Neutral**: SOTU speech that does not explicitly express support or disapproval of a public health program or initiative. This also includes speeches that do not explicitly/specifically reference public or private healthcare, or healthcare at all.
-   **Disapproval**: SOTU speech expressing dissatisfaction or highlighting shortcomings, failures, or barriers related to a public health initiative. This can also take the form of having approval for private health initiatives (i.e. support for the privatization of health care programs).

Additionally, we created a code book with these definitions to further clarify these categories. The code book also includes some keywords and example quotes from SOTU speeches for reference.
```

| **Sentiment** | **Definition** | **Keywords** | **Example** |
|------------------|------------------|------------------|-------------------|
| Approval | SOTU speech expressing satisfaction or belief in the value of a public health program or initiative. This can also take the form of having disapproval for private health initiatives (i.e. lack of support of the privatization of health care programs) | "support", "advocate", "commitment", "achieve", "improve", "success", "priority", "benefit", "impact", "significant progress", "solution", "progress", "invest", "improved health", "efforts have paid off" -- note: negation in conjunction with these words would be associated with disapproval | "Our Government has a responsibility to provide health care for the poor and the elderly, and we are meeting that responsibility. For all Americans, we must confront the rising cost of care, strengthen the doctor-patient relationship, and help people afford the insurance coverage they need." |
| Neutral | SOTU speech that does not explicitly express support or disapproval of a public health program or initiative. This also includes speeches that do not explicitly/specifically reference public or private healthcare, or healthcare at all. | "address", "discuss", "work", "plan", "explore", "next steps", "looking forward", "consider" | "I see a new America as we celebrate our 200th anniversary 6 years from now. I see an America in which we have abolished hunger, provided the means for every family in the Nation to obtain a minimum income, made enormous progress in providing better housing, faster transportation, improved health, and superior education." |
| Disapproval | SOTU speech expressing dissatisfaction or highlighting shortcomings, failures, or barriers related to a public health initiative. This can also take the form of having approval for private health initiatives (i.e. support for the privatization of health care programs) | "fail", "struggle", "insufficient", "lack", "decline", "not working", "unmet", "challenges", "shortcomings", "neglect", "ineffective", "difficult", "hurdle", "cut", "undermining", "deficient", etc. -- note: negation in conjunction with these words would be associated with approval | "In all we do, we must remember that the best health care decisions are not made by government and insurance companies but by patients and their doctors." |

After creating this code book, we selected a random sample of 80 documents from the SOTU corpus for hand coding. Out of the 80 speeches, half of these speeches were from Republicans and half were from Democrats; the speeches ranged from 1905 to 2013. Using this sample, two members of the group independently coded the 80 speeches according to the handbook and their interpretation of the text. The hand coded data was read back into R, and we computed confusion matrices and Krippendorff's alpha for all three categories to assess intercoder reliability.

-   

    ```{r}
    # random sample of 50
    set.seed(54321)
    random_50 <- sample_n(filtered_sotu, 50)

    # taking another sample of 30
    random_30 <- filtered_sotu %>%
      anti_join(random_50) %>%
      sample_n(30)

    # concatenate the two samples together
    sampled_df <- rbind(random_50, random_30)

    # get num of Republican speeches in the sample
    rep <- filter(sampled_df, party == 'Republican')

    # read handcoded results back into R
    handcode <- read.csv('handcoded.csv')

    sampled_df$Disapproval_1 <- handcode$Disapproval_1
    sampled_df$Neutral_1 <- handcode$Neutral_1
    sampled_df$Approval_1 <- handcode$Approval_1
    sampled_df$Disapproval_2 <- handcode$Disapproval_2
    sampled_df$Neutral_2 <- handcode$Neutral_2
    sampled_df$Approval_2 <- handcode$Approval_2

    # Disapproval Confusion Matrix
    table(sampled_df$Disapproval_1, sampled_df$Disapproval_2)
    # Krippendorff's alpha - 0.862 
    kripp.alpha(t(sampled_df[,c("Disapproval_1", "Disapproval_2")]))

    # Approval Confusion Matrix
    table(sampled_df$Approval_1, sampled_df$Approval_2)
    # Krippendorff's alpha - 0.82 
    kripp.alpha(t(sampled_df[,c("Approval_1", "Approval_2")]))

    # Neutral Confusion Matrix
    table(sampled_df$Neutral_1, sampled_df$Neutral_2)
    # Krippendorff's alpha - 0.876 
    kripp.alpha(t(sampled_df[,c("Neutral_1", "Neutral_2")]))

    ```

We had the highest intercoder reliability in regards to the Netural category, with a Krippendorff's alpha of 0.876. The second highest intercoder reliability was with the Dissapproval category, with a Krippendorff's alpha of 0.862. Our lowest intercoder reliability was with the Approval category, which had a Krippendorff's alpha of 0.82. From these metrics, it seems that our codebook captures and defines Disapproval and Neutral sentiments towards of public health care relatively well; however, there is some ambiguity in the codebook with regards to Approval sentiments.

-   Where did you codes conflict? \[1 paragraph\]

Across all three categories, our codes conflict the most in the Approval category, where we disagree on the coding of 7 out of 80 documents. For the Neutral category, we had 5 conflicts, where we differed on whether the speech should be coded as “Approval” or “Neutral.” This suggests that handcoding conflicts were more common when the speaker’s stance on public healthcare was ambiguous or nuanced, making it challenging to classify. Finally, the Disapproval category had the fewest conflicts, with disagreements on only 2 of 80 documents. This could indicate that disapproval is usually expressed in a clearer and more definitive manner, leaving less room for ambiguity in coding.

-   How would you revise your codebook based on these conflicts? \[1-2 paragraphs\]

Based on these conflicts, we would revise the definitions of each category to be clearer and more specific, especially the Approval and Neutral categories, since most of the disagreements derived from differentiating them. This could include adding decision rules regarding how to handle boundary cases in which the speaker might express tentative levels of support, or support while also encouraging caution. Further, the codebook can also be improved by providing examples of what counts as a public healthcare initiative. For example, we could clarify whether the definition should be limited to strictly government-funded healthcare programs or broadened to include any form of government influence in healthcare.

## Measurement

A: Use your hand coded data (you may need to hand code more training data) from step 1 to train a classifier to predict the categories you are interested in. Why did you choose the algorithm you did? \[1-2 paragraphs\] Use cross-validation to decide between variants of the models. Assess and report your out of sample precision and recall. \[1-2 paragraphs + table with precision and recall\]

Using our hand-coded dataset, we trained a Naive Bayes classifier to predict whether a SOTU speech expresses support for public health care. We selected this algorithm due to its ease of implementation, fast training time, and minimal computational requirements compared to more resource-intensive methods such as large language models. Additionally, the criteria outlined in our codebook for identifying support are relatively straightforward, making a simpler model well-suited to the task. While we acknowledge that Naive Bayes may struggle with more nuanced linguistic features such as negation, we believe it offers a reasonable balance between interpretability, efficiency, and predictive performance.

To train the Naive Bayes classifier, we first created a binary variable called `approval` in our dataset, which represents approval or non-approval (which includes both neutral or disapproval) of public health care initiatives. This variable is coded as $1$ if one or both coders categorize the speech as approval. If both coders label a speech as neutral or disapproval, this variables is coded as $0$. We then implemented a 75/25 train-validation split so that we may be able to assess out-of-sample fit. Of our 80 handcoded speeches, 60 speeches were used to train the classifier and 20 speeches were held out of training for validation.

```{r}
### Preprocessing dataset
# Create labels
sampled_df$approval <- ifelse(sampled_df$Approval_1==1 | sampled_df$Approval_2==1, 1, 0)
table(sampled_df$approval)

#head(sampled_df)
```

```{r}
table(sampled_df$approval)
# Note: approval/disapproval is slightly imbalanced
# This a binary label (approval = 1, neutral or disapproval = 0)
```

```{r}
#Preprocess the data
corpus_sotu <- corpus(sampled_df, text_field = "text")

#Create a document feature matrix (dfm)
#Some common pre-processing
toks <- tokens(corpus_sotu, remove_punct = TRUE, remove_numbers=TRUE)
toks <- tokens_wordstem(toks)
toks <- tokens_select(toks,  stopwords("en"), selection = "remove")
dfm <- dfm(toks)

#Split into training, validation, and unlabeled data
#Create an ID variable
docvars(corpus_sotu, "id_numeric") <- 1:ndoc(corpus_sotu)
#Unlabeled data
unlabeled <- which(is.na(sampled_df$approval))
#Labeled data
labeled <- which(!is.na(sampled_df$approval))
```

```{r}
#Sample training and validation set from the labeled data
set.seed(1234)
training <- sample(labeled, round(length(labeled)*.75))
length(training)
validation <- labeled[!labeled%in%training]
length(validation)

#Create separate dfm's for each
dfmat_train <- dfm_subset(dfm, docvars(corpus_sotu, "id_numeric")%in%training)
dfmat_val <- dfm_subset(dfm, docvars(corpus_sotu, "id_numeric")%in%validation)
```

```{r}
#Naive Bayes
#Train classifier
tmod_nb <- textmodel_nb(dfmat_train, docvars(dfmat_train, "approval"))
coef_nb <- coef(tmod_nb)
```

```{r}
#How well does it do in sample? - on training data
predict.train <- predict(tmod_nb, dfmat_train)
tab_train <- table(docvars(dfmat_train, "approval"), predict.train)
tab_train

#precision
diag(tab_train)/colSums(tab_train)
#recall
diag(tab_train)/rowSums(tab_train)
```

```{r}
#How well does this prediction do out of sample?  Validation
predict.val <- predict(tmod_nb, newdata = dfmat_val)
tab_val <- table(docvars(dfmat_val, "approval"), predict.val)
tab_val
#precision
diag(tab_val)/colSums(tab_val)
#recall
diag(tab_val)/rowSums(tab_val)
```

In regards to **in-sample** precision and recall, the Naive Bayes model does relatively well:

|           | 0 (neutral/disapproval) | 1 (approval) |
|-----------|-------------------------|--------------|
| precision | 0.912                   | 0.962        |
| recall    | 0.969                   | 0.893        |

However, it is more important to evaluate the Naive Bayes performance based on **out of sample** metrics.

|           | 0 (neutral/disapproval) | 1 (approval) |
|-----------|-------------------------|--------------|
| precision | 0.625                   | 0.417        |
| recall    | 0.417                   | 0.625        |

Based on these metrics, the Naive Bayes doesn't do extremely well. Based on the precision values, we can conclude that, of all predictions of neutral or disapproving sentiment towards public health care (0), 5 of the 8 were correct (0.625). Likewise, of all predictions of approving sentiment towards public health care (1), 5 of the 12 were correct (0.417). In regards to the recall values, we conclude that, of all actual handcoded examples of neutral/disapproving sentiment towards public health care, the model only predicted a proportion of 0.417 correctly. For all actual handcoded examples of approving sentiment, the model only predicted a proportion of 0.625 correctly. In short, our model is predicting neutral/disapproval more conservatively, yet more accurately, while predicting approval more often, with lower accuracy. The model overall has a relatively weak performance with none of the classes yielding thresholds above 0.70. Other classifier methods could be tested out of sample and compared to see if they yield better results.

```{r}
#Apply Naive Bayes to the handcoded data:
sampled_df$predict.approval <- as.numeric(as.character(predict(tmod_nb, newdata = dfm)))

#Here is the prediction of approval for the whole corpus
prop.table(table(sampled_df$predict.approval))

# If handcoding is representative of whole corpus,
# Slightly underpredicting non-approval, slightly overpredicting approval
prop.table(table(sampled_df$approval))
```

Apply either A or B to the whole corpus. Use the measure to answer your research question. \[1-2 paragraphs plus plot\] What do you learn? \[1-2 paragraphs\] What are some of the limitations of this approach? \[1-2 paragraphs\]

```{r}
#Preprocess the data
corpus_sotu_all <- corpus(filtered_sotu, text_field = "text")

#Create a document feature matrix (dfm)
#Some common pre-processing
toks_all <- tokens(corpus_sotu_all, remove_punct = TRUE, remove_numbers=TRUE)
toks_all <- tokens_wordstem(toks_all)
toks_all <- tokens_select(toks_all,  stopwords("en"), selection = "remove")
dfm_all <- dfm(toks_all)

# Match features in new DFM to the ones used in training
dfm_all_matched <- dfm_match(dfm_all, features = featnames(tmod_nb))

# Make predictions
predictions <- predict(tmod_nb, newdata = dfm_all_matched)

# Attach predictions to original data
filtered_sotu$predicted_sentiment <- predictions

sum(filtered_sotu$predicted_sentiment == 1)
```

```{r}
# Step 2: Count and compute proportions
plot_data <- filtered_sotu %>%
  mutate(predicted_sentiment = as.factor(predicted_sentiment)) %>%
  group_by(party, predicted_sentiment) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(party) %>%
  mutate(percentage = count / sum(count))

# Step 3: Plot percent stacked bar chart
ggplot(plot_data, aes(x = party, y = percentage, fill = predicted_sentiment)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) + scale_fill_manual(
    values = c("0" = "#ff5a5a",  # Blue
               "1" = "#77b255")  # Red
  )  # or "Paired", "Dark2", "Pastel1", etc.
  labs(
    title = "Predicted Sentiment by Political Party",
    x = "Political Party",
    y = "Percentage of Speeches",
    fill = "Predicted Sentiment"
  ) +
  theme_minimal()
```

```{r}
# Count how many Democratic speeches
sum(filtered_sotu$party == "Democratic")

# Count how many Republican speeches
sum(filtered_sotu$party == "Republican")

# Two-proportion z-test
prop.test(c(37,25), c(61,64), correct = FALSE)
```

Our group chose to apply approach A, which is to train a Naive Bayes classifier model on our hand-coded data, on a set of 80 randomly sampled State of the Union address data spanning from 1903 to 2013. We aimed to answer the question: "Are there systematic differences in how Democratic and Republican presidents frame public health issues in terms of sentiment (approval vs. neutral vs. disapproval) in State of the Union Addresses?". Based on our classifier model, we can say that there likely are differences in how Democratic and Republican presidents have addressed healthcare in their SOTU addresses, but there is a need for future work due to the fact that our classifier did not have strong prediction accuracy. Although the classifier demonstrated limited overall accuracy, it still identified patterns in the text that aligned with our hand-coded labels.

One of the lessons we learned were that simpler classification models, such as the Naive Bayes, can fail to capture nuanced political language, such as rhetorical shifts, humor, and ambiguity. These elements, which are commonly present in political speeches, can made the underlying sentiment or message about public health difficult to pinpoint. Another lesson we learned is that the size of the available data play a crucial role in a model's prediction accuracy.

Because we used a Naive Bayes classifier, we face several important limitations. The first is that Naive Bayes largely operates on the assumption that all predictors are independent of each other, which is a circumstance that rarely occurs in real life. In the case of our classifier, treating words as predictors can be problematic, as words in natural language are often highly interdependent. Another limitation is that the Naive Bayes model struggles with more nuanced linguistic features, such as negation. As discussed previously, political speech is often complex, therefore important information about how the speaker communicates about public health can be missed.

**References:**

Copeland, L. (2025, May 22). *House passes bill that would devastate health coverage for millions in Medicaid and Medicare*. Medicare Rights Center. https://www.medicarerights.org/medicare-watch/2025/05/22/house-passes-bill-that-would-devastate-health-coverage-for-millions-in-medicaid-and-medicare

*History, Evolution, and Practices of the President’s State of the Union Address: Frequently asked questions*. (n.d.). Congress.Gov \| Library of Congress. https://www.congress.gov/crs-product/R44770

Rutter, T. (1998). Health Features Prominently in the State of the Union Address. *BMJ: British Medical Journal*, *316*(7129), 415–415. http://www.jstor.org/stable/25176997

Teten, R. L. (2003). Evolution of the Modern Rhetorical Presidency: Presidential Presentation and Development of the State of the Union Address. *Presidential Studies Quarterly*, *33*(2), 333–346. http://www.jstor.org/stable/27552488
