---
title: "SOTU Analysis - POLI 176 Final Project"
format: html
editor: visual
---

# 1.

```{r}
sotu <- read.csv("SOTU_WithText.csv")

# filter from 1900s on 
filtered_sotu <- filter(sotu, year >= 1900)

set.seed(123123)
sample_n(filtered_sotu, 2)

```

```{r}
set.seed(12345)
sample_n(filtered_sotu, 2)
```

Select a corpus to study. Do some background reading on the corpus and decide on a research question that you can answer with the data that you have. What is your research question? What is the significance of your research question? Do a quick search -- has anyone studied this question before? (Please include at least 2 references) \[2-3 paragraphs\]

# 2.

Complete the steps below:

Step 1: Hand Coding

```         
- Take random sample of at least 50, code whether the document is for, against, or neutral.
```

-   Define a categorization that would help you answer your research question. \[1-2 paragraphs\]

    -   public food assistance
    -   education
    -   public healthcare assistance

    Categories:

    -   favor
    -   not in favor
    -   neutral

-   Create a detailed codebook. \[Include codebook\]

    -   We need to figure out what metrics we are categorizing on:
    -   sentiment words

-   Select a random sample from your data of at least 50 documents for hand coding. Have at least two people code the data, read the data back into R, and assess inter-coder reliability. \[1-2 paragraphs plus metrics on intercoder reliability\]

-   Where did you codes conflict? \[1 paragraph\]

-   How would you revise your codebook based on these conflicts? \[1-2 paragraphs\]

Step 2: Measurement

(Choose A or B or both).

A: Use your hand coded data (you may need to hand code more training data) from step 1 to train a classifier to predict the categories you are interested in. Why did you choose the algorithm you did? \[1-2 paragraphs\] Use cross-validation to decide between variants of the models. Assess and report your out of sample precision and recall. \[1-2 paragraphs + table with precision and recall\]

B: Use a large language model to predict the categories instead. What prompt did you choose and why? \[1-2 paragraphs\] Use your hand coded data to assess and report your out of sample precision and recall. \[1-2 paragraphs + table with precision and recall\]

Apply either A or B to the whole corpus. Use the measure to answer your research question. \[1-2 paragraphs plus plot\] What do you learn? \[1-2 paragraphs\] What are some of the limitations of this approach? \[1-2 paragraphs\]

**References:**

Please include a works cited or reference list at the end of your memo, with in-text references to any other work that you reference throughout.All work should be properly cited and the written work should be your own -- not taken from another source or generated by AI.
